\chapter{Semantics}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Semantics of Program}
\label{sec:semantics}

Semantics of the program defines the \textbf{meaning} of the program
that is grammatically correct.

\subsection{Prerequisites}

\subsubsection{Simple Language Syntax}

\begin{math}
  \begin{array}{llll}
    a \to & \mid n & \mid x & \mid a_1 + a_2 \\
          & \mid a_1 \times a_2 & \mid a_1 - a_2 \\
    b \to & \mid \mathtt{true} & \mid \mathtt{false} & \mid a_1 = a_2  \\
          & \mid a_1 \leq a_2 & \mid \neg b & \mid b_1 \land b_2 \\
    c \to & \mid x := a  & \mid \mathtt{skip} & \mid c_1 ; c_2 \\
          & \mid \mathtt{if} \text{ }b \text{ }c_1\text{ }c_2 & \mid \mathtt{while} \text{ }b \text{ }c \\
  \end{array}
\end{math}

A state $ s \in State $ is a function from variables to values:
\begin{math}
  \begin{array}{rcl}
    State & : & Var \to \mathbb{Z} \\
  \end{array}
\end{math}


\subsubsection{Semantics of Arithmetic Expressions}

\begin{math}
  \begin{array}{rcl}
    \mathcal{A} & : & \mathtt{Aexp} \to State \to \mathbb{Z} \\
    \mathcal{A} \SEM{ a } & : & State \to \mathbb{Z} \\
    \mathcal{A} \SEM{ n } (s) & = & n \\
    \mathcal{A} \SEM{ x } (s) & = & s(x) \\
    \mathcal{A} \SEM{ a_1 + a_2 } (s) & = & \mathcal{A} \SEM{ a_1 } (s) + \mathcal{A} \SEM{ a_2 } (s) \\
    \mathcal{A} \SEM{ a_1 \times a_2 } (s) & = & \mathcal{A} \SEM{ a_1 } (s) \times \mathcal{A} \SEM{ a_2 } (s) \\
    \mathcal{A} \SEM{ a_1 - a_2 } (s) & = & \mathcal{A} \SEM{ a_1 } (s) - \mathcal{A} \SEM{ a_2 } (s) \\
  \end{array}
\end{math}

\subsubsection{Semantics of Boolean Expressions}

\begin{math}
  \begin{array}{rcl}
    \mathcal{B} & : & \mathtt{Bexp} \to State \to {T} \\
    {T} & = & \{ {true}, {false} \} \\
    \mathcal{B} \SEM{ b } & : & State \to {T} \\
    \mathcal{B} \SEM{ \mathtt{true} } (s) & = & {true} \\
    \mathcal{B} \SEM{ \mathtt{false} } (s) & = & {false} \\
    \mathcal{B} \SEM{ a_1 = a_2 } (s) & = & \mathcal{A} \SEM{ a_1 } (s) = \mathcal{A} \SEM{ a_2 } (s) \\
    \mathcal{B} \SEM{ a_1 \leq a_2 } (s) & = & \mathcal{A} \SEM{ a_1 } (s) \leq \mathcal{A} \SEM{ a_2 } (s) \\
    \mathcal{B} \SEM{ \neg b } (s) & = & \mathcal{B} \SEM{ b } (s) = {false} \\
    \mathcal{B} \SEM{ b_1 \land b_2 } (s) & = & \mathcal{B} \SEM{ b_1 } (s) \wedge \mathcal{B} \SEM{ b_2 } (s) \\
  \end{array}
\end{math}

\subsubsection{Free Variables}

A set of variables occurring in the expression.

\begin{math}
  \begin{array}{rcl}
    FV(n) & = & \emptyset \\
    FV(x) & = & \{ x \} \\
    FV(a_1 + a_2) & = & FV(a_1) \cup FV(a_2) \\
    FV(a_1 \times a_2) & = & FV(a_1) \cup FV(a_2) \\
    FV(a_1 - a_2) & = & FV(a_1) \cup FV(a_2) \\
  \end{array}
\end{math}

q\paragraph{Lemma}

Let $ s $ and $ s' $ be two states satisfying that $ s(x) = s'(x) $
for all $ x \in FV(a) $.

Then, $ \mathcal{A} \SEM{ a } (s) =\mathcal{A} \SEM{ a } (s') $ holds.

\paragraph{Lemma}

Let $ s $ and $ s' $ be two states satisrying that $ s(x) = s'(x) $
for all $ x \in FV(b) $.

Then, $ \mathcal{B} \SEM{ b } (s) = \mathcal{B} \SEM{ b } (s') $
holds.


\subsubsection{Substitution}

$ a [ y \mapsto a_0 ] $ means the arithmetic expression obtained by
replacing each occurrence of $ y $ in $a $ by $ a_0 $.

\begin{math}
  \begin{array}{rcl}
    n[y \mapsto a_0 ] & = & n \\
    x[y \mapsto a_0 ] & = & \begin{cases}
      a_0 & \mathtt{if} x = y \\
      x & \mathtt{if} x \neq y \\
    \end{cases} \\
    (a_1 + a_2)[y \mapsto a_0] & = & (a_1[y \mapsto a_0]) + (a_2[y \mapsto a_0]) \\
    (a_1 \times a_2)[y \mapsto a_0] & = & (a_1[y \mapsto a_0]) \times (a_2[y \mapsto a_0]) \\
    (a_1 - a_2)[y \mapsto a_0] & = & (a_1[y \mapsto a_0]) - (a_2[y \mapsto a_0]) \\
  \end{array}
\end{math}


\begin{math}
  (s[y \mapsto v])(x) =
  \begin{cases}
    v & \mathtt{if} x = y \\
    s(x) & \mathtt{if} x \neq y \\
  \end{cases}
\end{math}



\paragraph{Lemma}
$\mathcal{A} \SEM{ a[y \mapsto a_0] } (s) = \mathcal{A} \SEM{a} (s [ y
\mapsto \mathcal{A} \SEM{a_0} (s)]) $ for alll state s.



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Operational Semantics}

Also called as \textit{transitional} semantics.

Semantics of a program is defined by the computation steps executed on
a machine. In other words, operational semantics is concerned about
\textbf{how to execute} the program and not merly what the execution
result is.

The semantics is defined as a transition system
$ ( \mathbb{S}, \to ) $, where

\begin{itemize}
\item $ \mathbb{S} $ is the set of all \textbf{possible states}.
\item $ \to \subseteq \mathbb{S} \times \mathbb{S} $ is a
  \textbf{transition relation} between two states. Describes how the
  execution takes place.
\item $ s \in \mathbb{S} $ is a \textbf{state} of the program. It is
  either $ \pair{ S, s }$ which is a non-terminal state (i.e., the
  statement $ S $ is to be executed from the state $ s $), or $ s $
  which is a terminal state.
\end{itemize}

There are two approaches for operational semantics.  The difference
between the two are in the definitions of \textsl{transition
  relation}.

\subsubsection{Big-step}
\label{sec:big-step}

Big-step operational semantics describes how the \textbf{overall
  results} of executions are obtained.

The transition relation specified the relationship between the initial
state and the final state: $ \pair{ S, s } \to s' $

Transition relation is defined with \textbf{inference rules} of the
form:
\\

\begin{math}
\frac{ \pair{ S_1, s_1} \to s_1', ..., \pair{ S_n, s_n} \to s_n' }
{ \pair{S, s} \to s' }
\text{ if ... }
\end{math}

where

\begin{itemize}
\item $ S_1, ..., S_n $ are \textsl{statements} that consistute $ S $.
\item A rule has a number of premises and one conclusion.
\item A rule may also have a number of conditions that have to be
  fulfilled whenever the rule is applied.
\item Rules without premises are called \textbf{axioms}.
\end{itemize}

Big-step operational semantics for the while language is:


\begin{tabularx}{\textwidth}{c}
\\
$ \overline{ \pair{ x := a, s } \to s [ x \mapsto \mathcal{A}\SEM{a}(s) ] } $\\
\\
$ \overline{ \pair{ \mathtt{skip}, s} \to s } $\\
\\

$ \underline{ \pair{S_1, s} \to s' \text{   } \pair{S_2, s'} \to s''} $ \\
$ \pair{S_1;S_2, s} \to s'' $ \\
\\

\end{tabularx}


\paragraph{Semantic Function for Statements}

The \textbf{semantic function} for statements is the partial function:

\begin{math}
  \begin{array}{rcl}
    \mathcal{S}_b & : & Stm \to ( State \to State) \\
    \mathcal{S}_b \SEM{S} (s) & = &
                                    \begin{cases}
                                      s' & \mathtt{if} \pair{S, s} \to s' \\
                                      \mathbf{undef} & \mathtt{otherwise}
                                    \end{cases}

  \end{array}
\end{math}


\subsubsection{Small-step}
\label{sec:small-step}

Small-step operational semantics describes how the \textbf{individual
  steps} of the computations take place.

The individual computation steps are described by the transition
relation of the form:
\\

\begin{math}
  \pair{S, s} \implies \gamma
\end{math}

where $\gamma$ is either non-terminal state or terminal state. The
transition expresses the first step of the execution of $S$ from state
$s$.

\begin{itemize}
\item If $ \gamma = \pair{S', s'} $, non-terminal state, then the
  execution of $S$ from $s$ is not completed and the remaining
  computation continues with $\pair{S', s'}$.
\item If $ \gamma = s'$, terminal state, then the execution of $S$
  from $s$ has terminated and the final state is $s'$.
\end{itemize}

We say $\pair{S, s}$ is \textbf{stuck} if there is no $\gamma$ such
that $\pair{S, s} \implies \gamma$.

\paragraph{Semantic Function}

The semantic function $\mathcal{S}_s$ for small-step semantics is:
\begin{math}
  \begin{array}{rcl}
    \mathcal{S}_s & : & Stm \to (State \to State) \\
    \mathcal{S}_s \SEM{S} (s) & = &
                                    \begin{cases}
                                      s' & \mathtt{if} \pair{S, s} \xRightarrow{*} s' \\
                                      \mathbf{undef} & \mathtt{otherwise}
                                    \end{cases}

  \end{array}
\end{math}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Denotational Semantics}
Also called as \textit{compositional} semantics.

Semantics of a program is defined by the semantics of the sub-parts of
the program. Thus, proving its soundness is by structural induction on
the program. For some realistic programming languages, even defining
their compositional semantics is an obstacle, due to \texttt{goto},
\texttt{exception}, or \texttt{call}.


%%% Local Variables:
%%% mode: latex
%%% TeX-master: "program-analysis"
%%% End:
